---
title: "Data Prevalidation Preparation"
output: html_notebook
---

## 1. Google Earth Engine

GEE images code link: <https://code.earthengine.google.com/3d798edab81ae80002b71aa8a58ee5db>

GEE images count code link: <https://code.earthengine.google.com/e82b9d4616df0c3b05dd6f7c280cdc1d>

**Data Available:**

-   2022: Landsat 8

-   2024: Landsat 8 and 9 (merge together?)

-   2025: Landsat 9

## 2. r5r

### 2.1. Installation

**Install r5r:**

```{r}
# Code Source: https://ipeagit.github.io/r5r/
# from CRAN
# install.packages("r5r")

# dev version with latest features
# utils::remove.packages('r5r')
# devtools::install_github("ipeaGIT/r5r", subdir = "r-package")
```

**Install JDK using {rJavaEnv} package in order to use r5r:**

```         
Downloaded in: C:\Users\choic\AppData\Local\R\cache\R\rJavaEnv\installed\windows\x64\21
Also downloaded in: C:\Users\choic\CASA\Dissertation_Data\rjavaenv\windows\x64\21
```

```{r}
# install.packages('rJavaEnv')

# check version of Java currently installed (if any) 
rJavaEnv::java_check_version_rjava()

# install Java 21
rJavaEnv::java_quick_install(version = 21)
```

**Load r5r:**

```{r}
# Install from CRAN
# install.packages("r5r")

# Warning message: Currently, Java memory is set to "-Xmx512m"
# Warning message suggest set Java max memory to 2 GB before loading r5r
options(java.parameters = '-Xmx2G')
library(r5r)
```

### 2.2. Data Sources

**Data requirements for r5r :** <https://ipeagit.github.io/r5r/#data-requirements>

-   Road network data set from OpenStreetMap in .pbf format

-   Public transport feed in GTFS.zip format

+-----------------------+---------------------------------------------------------------------------------------------------------------------------------------+---------------------------------------------------------------------------------------+
| Data                  | Description                                                                                                                           | Link                                                                                  |
+=======================+=======================================================================================================================================+=======================================================================================+
| GTFS                  | Download as NaPTAN file                                                                                                               | <https://beta-naptan.dft.gov.uk/>                                                     |
|                       |                                                                                                                                       |                                                                                       |
|                       | It then needs to be converted into a GTFS format (<https://citygeographics.org/r5r-workshop/uk-transit-data-transxchange-and-atoc/>). |                                                                                       |
+-----------------------+---------------------------------------------------------------------------------------------------------------------------------------+---------------------------------------------------------------------------------------+
| GTFS                  | Download L.zip (for London) (downloaded on 16/04/2026)                                                                                | Look for Core FTP and use the access host, username and password from the sent email. |
+-----------------------+---------------------------------------------------------------------------------------------------------------------------------------+---------------------------------------------------------------------------------------+
| Road Network from OSM | Downloaded as .pbf file                                                                                                               | <https://download.geofabrik.de/europe/united-kingdom/england/greater-london.html>     |
+-----------------------+---------------------------------------------------------------------------------------------------------------------------------------+---------------------------------------------------------------------------------------+

### 2.2. GTFS Data 

#### 2.2.1. Convert data into GTFS format

```{r, eval=FALSE}
remotes::install_github("ITSleeds/UK2GTFS")
```

```{r, eval=FALSE}
# Convert London TNDS (TransXChange) data into a GTFS object for later use in r5r

# Load the specific package for converting to GTFS
library(UK2GTFS)

# Load Packages
library(readr)
library(dplyr)


# Read NaPTAN stop data and rename columns for use with UK2GTFS
naptan_local <- read_csv("r5r_london/Stops.csv", show_col_types = FALSE) %>%
  rename(
    stop_id = ATCOCode,
    stop_name = CommonName,
    stop_lat = Latitude,
    stop_lon = Longitude,
    stop_type = StopType
  )

# %>%
#    filter(!is.na(stop_id),
#          !is.na(stop_name),
#          !is.na(stop_lat),
#          !is.na(stop_lon))



# Using the L.zip (London specific file) and convert into a GTFS object
path_in <- "r5r_london/L.zip"
gtfs <- transxchange2gtfs(path_in = path_in,
                          # remove this line if it is not local
                          # naptan = naptan_local,
                          ncores = 5)
```

```{r, eval=FALSE}
# Save as GTFS file
gtfs_write(gtfs, 
           folder = "r5r_london/",
           name = "gtfs_london")
```

#### GTFS

```{r}
# Load in GTFS file

# install.packages("gtfstools")
library(gtfstools)

gtfs <- read_gtfs("r5r_london/gtfs_london.zip")
```

```{r}
# Check
names(gtfs)
sapply(gtfs, nrow)
```

```{r}
# Check any missing data
length(setdiff(unique(gtfs$stop_times$trip_id), unique(gtfs$trips$trip_id)))
length(setdiff(unique(gtfs$trips$route_id), unique(gtfs$routes$route_id)))
length(setdiff(unique(gtfs$stop_times$stop_id), unique(gtfs$stops$stop_id)))

calendar_service_ids <- if ("calendar" %in% names(gtfs)) unique(gtfs$calendar$service_id) else character(0)
calendar_dates_service_ids <- if ("calendar_dates" %in% names(gtfs)) unique(gtfs$calendar_dates$service_id) else character(0)
length(setdiff(unique(gtfs$trips$service_id), union(calendar_service_ids, calendar_dates_service_ids)))
```

```{r}
range(gtfs$calendar$start_date)
range(gtfs$calendar$end_date)
```
