---
title: "r5r Data Preparation"
---

### 1. Installation

**Install r5r:**

```{r}
# Code Source: https://ipeagit.github.io/r5r/
# from CRAN
# install.packages("r5r")

# dev version with latest features
# utils::remove.packages('r5r')
# devtools::install_github("ipeaGIT/r5r", subdir = "r-package")
```

**Install JDK using {rJavaEnv} package in order to use r5r:**

```         
Java version: "21.0.10"
Installed at: C:\Users\choic\AppData\Local\R\cache\R\rJavaEnv\installed\windows\x64\21
Symlinked to: C:\Users\choic\CASA\Dissertation_Data\rjavaenv\windows\x64\21
```

```{r}
# install.packages('rJavaEnv')

# check version of Java currently installed (if any) 
rJavaEnv::java_check_version_rjava()

# install Java 21
rJavaEnv::java_quick_install(version = 21)
```

**Load r5r:**

```{r}
# Install from CRAN
# install.packages("r5r")

# Warning message: Currently, Java memory is set to "-Xmx512m"
# Warning message suggest set Java max memory to 2 GB before loading r5r
options(java.parameters = '-Xmx2G')
library(r5r)
```

### 2. Data

**Data requirements for r5r** <https://ipeagit.github.io/r5r/#data-requirements> **:**

-   Road network data set from OpenStreetMap in .pbf format

-   Public transport feed in GTFS.zip format

| Data | Description | Source |
|------------------|---------------------------------|---------------------|
| Stops.csv (Stops location) | Download as NaPTAN (National Public Transport Access Nodes) file, to be connected to TransXChange data, to be converted into a GTFS format (<https://citygeographics.org/r5r-workshop/uk-transit-data-transxchange-and-atoc/>). | <https://beta-naptan.dft.gov.uk/> |
| L.zip (TransXChange timetable data) | Download L.zip (for London) (downloaded on 16/04/2026), to be converted into GTFS | Look for Core FTP and use the access host, username and password from the sent email |
| timetable.zip (Train timetable data) | Convert to GTFS | <https://opendata.nationalrail.co.uk/> |
| OSM in .pbf format (Road network) | Downloaded as .pbf file, to be used directly for r5r | <https://download.geofabrik.de/europe/united-kingdom/england/greater-london.html> |

### 3. GTFS Data

#### 3.1. Bus, Tube, Tram Data

**Convert data into GTFS format**

```{r, eval=FALSE}
remotes::install_github("ITSleeds/UK2GTFS")
```

```{r, eval=FALSE}
# Convert London TNDS (TransXChange) data into a GTFS object for later use in r5r

# Load the specific package for converting to GTFS
library(UK2GTFS)

# Load Packages
library(readr)
library(dplyr)


# Read NaPTAN stop data and rename columns for use with UK2GTFS
naptan_local <- read_csv("r5r_data/raw_data/Stops.csv", show_col_types = FALSE) %>%
  rename(
    stop_id = ATCOCode,
    stop_name = CommonName,
    stop_lat = Latitude,
    stop_lon = Longitude,
    stop_type = StopType
  )

# %>%
#    filter(!is.na(stop_id),
#          !is.na(stop_name),
#          !is.na(stop_lat),
#          !is.na(stop_lon))



# Using the L.zip (London specific file) and convert into a GTFS object
path_in <- "r5r_data/raw_data/L.zip"
gtfs <- transxchange2gtfs(path_in = path_in,
                          # remove this line if it is not local
                          naptan = naptan_local,
                          ncores = 5)
```

```         
MISSING VALUES:
London Underground (9400ZZLU...) — 9400ZZLUBTK2, 9400ZZLUCND2, 9400ZZLUHCL2, 9400ZZLUBTX2, 9400ZZLUCFM2, 9400ZZLUMTC2, 9400ZZLUGDG2, 9400ZZLUWLO7, 9400ZZLUAGL2, 9400ZZLUODS2, 9400ZZLULNB4, 9400ZZLUCPN2, 9400ZZLUCPC2, 9400ZZLUCPS2, 9400ZZLUBLM2, 9400ZZLUTBC2, 9400ZZLUTBY2, 9400ZZLUCSD2, 9400ZZLUSWN2, 9400ZZLUHGT2, 9400ZZLUACY2, 9400ZZLUHWC2, 9400ZZLUHWT2, 9400ZZLUHNX2, 9400ZZLUSGT2, 9400ZZLUBDS2, 9400ZZLUWOG2, 9400ZZLUTPN2, 9400ZZLUMRH2, 9400ZZLUHWY2, 9400ZZLUGPK6, 9400ZZLUHPC2, 9400ZZLUCAR2, 9400ZZLUOAK2, 9400ZZLUSFS2, 9400ZZLUWIP2, 9400ZZLUGBY2, 9400ZZLUUPB2, 9400ZZLUEPK2, 9400ZZLUUPY2, 9400ZZLUGTH2, 9400ZZLURBG2, 9400ZZLUWSD2, 9400ZZLUBLG2, 9400ZZLUBND4, 9400ZZLUMBA2, 9400ZZLULGT2, 9400ZZLUQWY2, 9400ZZLUSBC2, 9400ZZLUHGR2, 9400ZZLUPVL2, 9400ZZLUNHT2, 9400ZZLUSRP2, 9400ZZLURSG2, 9400ZZLUCYF2, 9400ZZLULNB3, 9400ZZLUWLO8, 9400ZZLUWHP2, 9400ZZLUWIG2, 9400ZZLUSWC2, 9400ZZLUSJW2, 9400ZZLUNDN2, 9400ZZLUDOH2, 9400ZZLUKSX8, 9400ZZLUNKP2, 9400ZZLUPRD2, 9400ZZLUBLR2, 9400ZZLUTMH2, 9400ZZLUPCO2, 9400ZZLUVXL2, 9400ZZLUSKT2, 9400ZZLUKPK2, 9400ZZLUMVL2, 9400ZZLUWKA2, 9400ZZLULBN2, 9400ZZLURYO2
DLR (9400ZZDL...) — 9400ZZDLPOP4, 9400ZZDLSHA2, 9400ZZDLTWG2, 9400ZZDLKGV2, 9400ZZDLISL2, 9400ZZDLCUT2, 9400ZZDLBNK2, 9400ZZDLLCA2, 9400ZZDLPRE2, 9400ZZDLCUS2
Bus (490...) — 490000283N
```

**Save as GTFS data file**

```{r, eval=FALSE}
# Save as GTFS file
gtfs_write(gtfs, 
           folder = "r5r_data/r5r_london_data",
           name = "gtfs_london")
```

#### 3.2. Rail Data

```{r}
# Data taken from National Rail Portal
# Save username and password
install.packages("usethis")
library(usethis)

usethis::edit_r_environ()
```

```{r}
# Check if username is saved
# Sys.getenv("NRDP_username")
```

```{r}
# Restarted R session, so load this library again
# Load the specific package for converting to GTFS
library(UK2GTFS)

nrdp_timetable("r5r_data/raw_data/timetable.zip")

path_in <- "r5r_data/raw_data/timetable.zip"
gtfs_rail <- atoc2gtfs(path_in = path_in,
                       ncores = 5)

gtfs_write(gtfs_rail,
           folder = "r5r_data/r5r_london_data",
           name = "gtfs_rail"
)
```

Check what companies there are in the file (for filtering to London services only):

```{r}
library(UK2GTFS)
library(tidytransit)

# read your already converted rail GTFS
gtfs_rail <- read_gtfs("r5r_data/r5r_london_data/gtfs_rail.zip")

# see all agency names
gtfs_rail$agency %>%
  select(agency_id, agency_name) %>%
  arrange(agency_name)
```

```{r}
# see all codes you have
gtfs_rail$agency %>% pull(agency_id)
```

```{r}
# Select London railway services operators
# http://www.railwaycodes.org.uk/operators/toccodes.shtm
london_operators <- c(
   # Key London operators
  "LT",  # London Underground
  "LO",  # London Overground
  "XR",  # Elizabeth line
  
  # Wider London Operators
  "TL",  # Thameslink: St Pancras, City Thameslink, Blackfriars, London Bridge
  "GN",  # Great Northern: King's Cross, Moorgate/Old Street, ...north London
  "SN",  # Southern: Victoria, London Bridge, ...south London
  "SE",  # Southeastern: Charing Cross, Cannon Street, Victoria, London Bridge...
  "SW",  # South Western Railway: Waterloo + Vauxhall; south-west London
  "GX"   # Gatwick Express
)

# filter to only london operators
gtfs_rail_london <- gtfs_rail
gtfs_rail_london$agency <- gtfs_rail$agency %>%
  filter(agency_id %in% london_operators)

# filter routes
gtfs_rail_london$routes <- gtfs_rail$routes %>%
  filter(agency_id %in% london_operators)

# filter trips 
gtfs_rail_london$trips <- gtfs_rail$trips %>%
  filter(route_id %in% gtfs_rail_london$routes$route_id)

# filter stop_times 
gtfs_rail_london$stop_times <- gtfs_rail$stop_times %>%
  filter(trip_id %in% gtfs_rail_london$trips$trip_id)

# Check before and after filtering
nrow(gtfs_rail$stop_times)
nrow(gtfs_rail_london$stop_times)
```

```{r}
# save
# gtfs_write(gtfs_rail_london, folder = "r5r_data/r5r_london_data", name = "gtfs_rail_london")
```
