---
title: "GTFS Data Conversion Attempt 2"
---

## 1. Load Packages

```{r}
# Load packages
library(sf)
library(dplyr)
library(tidyverse)
library(tidytransit)
```

## 2. Data Loading and Pre-processing

```{r}
# Load London LSOA shapefile
lsoa_london <- st_read("lsoa_london/lsoa_london.shp")
```

```{r}
# filter to all Camden LSOAs
camden_lsoa <- lsoa_london %>%
  filter(grepl("Camden", LSOA21NM))

# check how many LSOAs Camden has
nrow(camden_lsoa)
```

```{r}
# Load Greenspace Access Points
greenspace_accesspoint_london <- st_read("OS-Open-Greenspace-GB/GB_AccessPoint/GB_AccessPoint.shp") %>%
  # Ensure consistent CRS
  st_transform(st_crs(lsoa_london)) %>%
  # Filter to London only
  .[st_intersects(., st_union(lsoa_london), sparse = FALSE), ]
```

## 3. Create origins and destinations

### Origins

```{r}
sf_use_s2(FALSE)

# create origin points
origins <- st_centroid(camden_lsoa) %>%
  # reproject
  st_transform(4326) %>%
  # create new column
  mutate(
    id  = LSOA21CD,
    lon = as.numeric(unlist(map(geometry,1))),
    lat = as.numeric(unlist(map(geometry,2)))
  )
# remove geometry column
origins$geometry <- NULL
# only keep id, lon and lat
origins <- origins %>%
  select(id, lon, lat)

head(origins)
```

```{r}
# Checks
nrow(origins) 
range(origins$lon) 
range(origins$lat)
sum(is.na(origins$lon)) 
```

### Destinations

Testing on Camden borough only, so filter green space access points to Camden only first:

```{r}
greenspace_accesspoint_camden <- greenspace_accesspoint_london[
  st_intersects(greenspace_accesspoint_london, st_union(camden_lsoa), sparse = FALSE), ]

nrow(greenspace_accesspoint_camden)
```

Set destinations now:

```{r}
destinations <- greenspace_accesspoint_camden %>%
  st_transform(4326) %>%
  mutate(
    id  = as.character(refToGSite),
    lon = as.numeric(unlist(map(geometry, 1))),
    lat = as.numeric(unlist(map(geometry, 2)))
  )

destinations$geometry <- NULL

destinations <- destinations %>%
  select(id, lon, lat)

head(destinations)
```

```{r}
# Checks
nrow(destinations) 
range(destinations$lon) 
range(destinations$lat)
sum(is.na(destinations$lon)) 
```

## 4. r5r

### GTFS data check

Check Bus Open Data Service GTFS route type and service providers:

```{r}
# gtfs_bus <- read_gtfs("r5r_data/r5r_london_data2/itm_london_gtfs.zip")

# Route types
# 3 = bus, 1 = subway/tube, 2 = rail, 0 = tram
table(gtfs_bus$routes$route_type)

# Service providers (agency)
unique(gtfs_bus$agency$agency_name)
```

Check rail GTFS:

Ok above does not seem to include overground. So read in rail gtfs to check if have it:

```{r}
#gtfs_rail <- read_gtfs("r5r_data/r5r_london_data/gtfs_rail.zip")

# Check all agencies
unique(gtfs_rail$agency$agency_name)
```

Check what TfL Rail includes (elizabeth line??):

```{r}
gtfs_rail$routes %>%
  left_join(gtfs_rail$agency, by = "agency_id") %>%
  filter(grepl("TfL Rail", agency_name)) %>%
  select(route_short_name, route_long_name, route_type)

# check lizzy line stops 
gtfs_rail$stops %>%
  filter(grepl("Abbey Wood|Shenfield|Paddington|Liverpool Street|Whitechapel|Canary Wharf|Custom House|Woolwich", 
               stop_name, ignore.case = TRUE)) %>%
  select(stop_name, stop_id) %>%
  arrange(stop_name)

# number of trips TfL Rail operates
gtfs_rail$trips %>%
  left_join(gtfs_rail$routes, by = "route_id") %>%
  left_join(gtfs_rail$agency, by = "agency_id") %>%
  filter(grepl("TfL Rail", agency_name)) %>%
  nrow()
```

### Build r5r network 

```{r}
# increase Java memory
options(java.parameters = "-Xmx12G")

# load libraries
library(r5r)

#path <- system.file("r5r_path", package = "r5r")
r5r_core <- r5r::build_network(data_path = "r5r_data/r5r_london_data2")

# used setup_r5() previously and got the following: 
# Warning: ! `setup_r5()` is being deprecated in *r5r* v2.3.0 and will be removed in a future release. ℹ Please use `r5r::build_network()` instead.
```

Check valid dates:

```{r}
library(tidytransit)

# Read each GTFS file and check calendar coverage
gtfs_bus  <- read_gtfs("r5r_data/r5r_london_data2/itm_london_gtfs.zip")
gtfs_rail <- read_gtfs("r5r_data/r5r_london_data2/gtfs_rail.zip")

# Check date ranges
cat("=== BUS GTFS ===\n")
cat("Start date:", as.character(min(gtfs_bus$calendar$start_date)), "\n")
cat("End date:  ", as.character(max(gtfs_bus$calendar$end_date)), "\n")

cat("\n=== RAIL GTFS ===\n")
cat("Start date:", as.character(min(gtfs_rail$calendar$start_date)), "\n")
cat("End date:  ", as.character(max(gtfs_rail$calendar$end_date)), "\n")
```

```{r}
# set routing paramaters

# maximum trip durations to be 1 hour (60 min)
max_trip_duration <- 60 

# can change to just a day (no time) and it will return min time of the entire day 
departure_datetime <- as.POSIXct("17-06-2026 08:30:00", 
                                 format = "%d-%m-%Y %H:%M:%S",
                                 tz = "Europe/London")
```

```{r}
# travel by public transport only
mode_transit <- ("TRANSIT")

# estimating the travel time
TTM_Camden_PT <- travel_time_matrix(r5r_core,
                          origins = origins,
                          destinations = destinations,
                          mode = mode_transit,
                          departure_datetime = departure_datetime,
                               max_trip_duration = max_trip_duration,
                          time_window = 120)
```

```{r}
nrow(TTM_Camden_PT)
head(TTM_Camden_PT, 20)
summary(TTM_Camden_PT$travel_time_p50)
```

```{r}
# travel by walking only 

mode_walk <- c("WALK")

# estimating the travel time
TTM_Camden_walk <- travel_time_matrix(r5r_core,
                          origins = origins,
                          destinations = destinations,
                          mode = mode_walk,
                          departure_datetime = departure_datetime,
                          max_trip_duration = max_trip_duration,
                          time_window = 120)
```

```{r}
nrow(TTM_Camden_walk)
head(TTM_Camden_walk, 20)
summary(TTM_Camden_walk$travel_time_p50)
```
